Fractional Types

### 2016-12-17 typespp.tex
fix compile error
a little on sec 3.7 but not done

For plain sets, the definition just counts the elements as each element is its
own equivalence class and has exactly one automorphism (the identity). Without
quite formalizing them and relying on the informal diagrams until the next
section, we argue that each of the groupoids $G_1$, $G_2$, and $G_3$ in
Fig.~\ref{fig:groupoids2} has cardinality $\frac{3}{2}$. Groupoid~$G_1$ consists
of two isomorphism classes: class~$a$ has one object with one automorphism (the
identity) and class~$c$ has one object with two distinct automorphisms; the
cardinality is $\frac{1}{1} + \frac{1}{2} = \frac{3}{2}$. For groupoid~$G_2$, we
also have two isomorphism classes with representatives $a$ and $c$; the class
containing $a$ has two automorphisms starting from $a$: the identity and the
loop going from $a$ to~$b$ and back. By the groupoid axioms, this loop is
equivalent to the identity which means that the class containing $a$ has just
one automorphism. The isomorphism class of $c$ has two non-equivalent
automorphisms on it and hence the cardinality of $G_2$ is also
$\frac{1}{1} + \frac{1}{2} = \frac{3}{2}$. For~$G_3$, we have three isomorphism
classes, each with two non-equivalent automorphisms, and hence the cardinality
of $G_3$ is $\frac{1}{2} + \frac{1}{2} + \frac{1}{2} = \frac{3}{2}$.
It is important to note that $G_1$ and $G_2$ are (categorically) equivalent
groupoids, but that $G_3$ is not equivalent to either $G_1$ or $G_2$.
Roughly speaking this is because the number of connected components is also
an invariant of a groupoid, and here $G_1$ and $G_2$ have $2$ whilst $G_3$ has
$3$.

@@

If we are to build a compositional programming language around iteration and
symmetry groupoids, we need to ensure that they compose sensibly with the
existing type formers. Groupoids, viewed as categories, come associated with
natural notions of sums and products, and one might expect or hope that
identities which hold for rational numbers lift to identities in our
situation. The most prominent and important identity we seek is:
\[
\order{p} \otimes \iorder{p} \simeq \mathbb{1}
\]
for some appropriate unit object $\mathbb{1}$. The groupoids on either side are not categorically equivalent but they do have the same cardinality. Furthermore the iteration groupoid and the symmetry groupoid naturally ``cancel out'' in the sense that the iteration groupoid introduces the iterates of a combinator $c$ as separate objects while the symmetry groupoid uses these same iterates to express internal symmetries. Thus if we focus on the right notion of morphisms, it is sensible to consider them equivalent. The situation is however subtle and does not directly support a notion of sums. It is studied in detail in the next section.

@@

at end of previous section; add a new subsection explaining functor showing
equivalence of 1 and #p x 1/#p. Morita functor. Pullback condition. there is a
choice for eta but there is an extra condition that the choices are all
consistent. The big question is how to encode this global condition; side
effect; synchronization; etc. so naturally lead to pointed types and revise types

>>>>

For plain sets, the definition just counts the elements as each element is its
own equivalence class and has exactly one automorphism (the identity). Without
quite formalizing them and relying on the informal diagrams until the next
section, we argue that each of the groupoids $G_1$, $G_2$, and $G_3$ in
Fig.~\ref{fig:groupoids2} has cardinality $\frac{3}{2}$. Groupoid~$G_1$ consists
of two isomorphism classes: class~$a$ has one object with one automorphism (the
identity) and class~$c$ has one object with two distinct automorphisms; the
cardinality is $\frac{1}{1} + \frac{1}{2} = \frac{3}{2}$. For groupoid~$G_2$, we
also have two isomorphism classes with representatives $a$ and $c$; the class
containing $a$ has two automorphisms starting from $a$: the identity and the
loop going from $a$ to~$b$ and back. By the groupoid axioms, this loop is
equivalent to the identity which means that the class containing $a$ has just
one automorphism. The isomorphism class of $c$ has two non-equivalent
automorphisms on it and hence the cardinality of $G_2$ is also
$\frac{1}{1} + \frac{1}{2} = \frac{3}{2}$. For~$G_3$, we have three isomorphism
classes, each with two non-equivalent automorphisms, and hence the cardinality
of $G_3$ is $\frac{1}{2} + \frac{1}{2} + \frac{1}{2} = \frac{3}{2}$.  It is
important to note that $G_1$ and $G_2$ are (categorically) equivalent groupoids,
but that $G_3$ is not equivalent to either $G_1$ or $G_2$.  Roughly speaking
this is because the number of connected components is also a categorical
invariant of a groupoid, and here $G_1$ and $G_2$ have $2$ whilst $G_3$ has $3$.

@@

If we are to build a compositional programming language around iteration and
symmetry groupoids, we need to ensure that they compose sensibly with the
existing type formers. Groupoids, viewed as categories, come associated with
natural notions of sums and products, and one might expect or hope that
identities which hold for rational numbers lift to identities in our
situation. The most prominent and important identity we seek is:
\[
\order{p} \otimes \iorder{p} \simeq \order{\idiso}
\]
The left hand side is the categorical product of the iteration groupoid and
symmetry groupoid for some arbitrary 1-combinator $p$. The cardinality of this
groupoid is 1. The right hand side is the iteration groupoid for the identity
1-combinator which also has cardinality 1. The groupoids on either side are not
categorically equivalent and hence $\simeq$ cannot be categorical equivalence.
There are however other notions of groupoid equivalence such as Morita
equivalence~\cite{looping} that identifies the two groupoids above.  Although
the notion of equivalence we use in the next section appears novel, it is
related ``in spirit'' to Morita equivalence and we find it useful to give a
small motivating example instantitating $p$ above to $\permfive$ (which has
order 3). Essentially we are verifying that the product of the groupoid
$\order{\permfive}$ of cardinality 3 and the groupoid $\iorder{\permfive}$ of
cardinality $\frac{1}{3}$ is the trivial groupoid with only one distinct object
with one trivial automorphism.

@@

at end of previous section; add a new subsection explaining functor showing
equivalence of 1 and $\order{p} \times \iorder{p}$. Morita functor. Pullback condition. there is a
choice for eta but there is an extra condition that the choices are all
consistent. The big question is how to encode this global condition; side
effect; synchronization; etc. so naturally lead to pointed types and revise types
